Parametricity for Primitive Nested Types and GADTs

### 2021-07-13 21-lmcs/21-lmcs.tex
MFPS revisions #9

Since the type systems of F$_\omega$ and dependent type theories are
designed to extend System F with far more than non-ADT data types, it
seems like serious overkill to pass to their parametric models to
reason about nested types in System F. Indeed, such calculi support
fundamentally new features --- e.g., the full hierarchy of
higher-kinds or term-indexed types --- adding complexity to their
models that is entirely unnecessary for reasoning about nested types.
In addition, to our knowledge no such models represent data types as
primitive fixpoints. (See the past paragraph of this section for a
more detailed discussion in the case of F$_\omega$.) Whether or not
the results of this paper even hold for such calculi is an open
question.

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]\\[3ex]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[5ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[6ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof\\[2ex]
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note, however, that
weakening does not change the contexts in which types can be formed.
For example, the functorial context in which a type $\Nat^\emptyset F
\,G$ is formed must be empty, although it can be weakened by a
functorial variable $\beta$ {\em after} the type is formed; by
contrast, we cannot {\em first} weaken the functorial contexts of $F$
and $G$ to $\Gamma; \beta \vdash F$ and $\Gamma; \beta \vdash G$, and
{\em then} form $\Nat^\emptyset F \,G$ in the weakened context
$\Gamma; \beta$.  If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset
\vdash G$, then our rules allow formation of the type
$\Gamma;\emptyset \vdash \Nat^\emptyset F \,G$, which represents the
System F type $\Gamma \vdash F \to G$. Similarly, if
$\Gamma;\ol{\alpha} \vdash F$, then our rules allow formation of the
type $\Gamma; \emptyset \vdash \Nat^{\ol\alpha} \,\onet \,F$, which
represents the System F type $\Gamma; \emptyset \vdash \forall
\ol\alpha . F$. However, some System F types, such as $\forall
\alpha. (\alpha \to \alpha) \to \alpha$, are not representable in our
calculus. Note that, in the rule for $\mu$-types, no variables in
$\Phi$ appear in the body $F$ of $\mu \phi. \lambda \ol\alpha.F$. This
will be critical to proving the Identity Extension Lemma for our
calculus.

@@

\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, G \,K\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
The idea is that the arguments to $\phi$ get substituted for the
variables in $\ol\alpha$ in each $F$ replacing an occurrence of
$\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$. {\color{red} REMOVE? In addition, if
  \,$\Gamma, \phi^k; \Phi \vdash H$,\, $\Gamma; \ol\psi,\ol{\alpha}
  \vdash F$, $|\ol\alpha| = k$, and $\Phi \cap \ol\psi = \emptyset$,
  then $\Gamma,\ol\psi';\Phi \vdash H[\phi :=_{\ol{\alpha}} F[\ol{\psi
        :== \psi'}]]$. Ask Daniel.}

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \emptyset
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \emptyset
         ~|~\emptyset
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha} \vdash H$}
       \UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
         \Nat^{\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha} \vdash H$}
       \AXC{$\Gamma; \ol\beta \vdash F$}
       \BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
         (\Nat^{\ol{\beta}}\,(\mu \phi.\lambda
         \ol{\alpha}.H)\ol{\beta}\,F)$} 
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

To define our term calculus we assume an infinite set $\cal V$ of term
variables disjoint from $\tvars$ and $\fvars$. If $\Gamma$ is a type
constructor context and $\Phi$ is a functorial context, then a {\em
  term context for $\Gamma$ and $\Phi$} is a finite set of bindings of
the form $x : F$, where $x \in {\cal V}$ and $\Gamma; \Phi \vdash
F$. We adopt the above conventions for denoting disjoint unions and
vectors in term contexts. If $\Delta$ is a term context for $\Gamma$
and $\Phi$, then the formation rules for the set of {\em well-formed
  terms over $\Delta$} are as in Figure~\ref{fig:terms}. In the rule
there for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all occurrences
of the type variables in $\ol{\alpha}$ in the types of $x$ and $t$, as
well as all occurrences of $x$ in $t$. In the rule for $t_{\ol K} s$
there is one type in $\ol K$ for every functorial
variable in $\ol \alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$
there is one type $F$ and one type
$G$ for each functorial variable in $\ol\phi$. Moreover, for each
$\phi^k$ in $\ol\phi$ the number of functorial variables in $\ol\beta$
in the judgments for its corresponding type $F$ and
$G$ is $k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
one $\beta$ for every $\alpha$. Substitution for terms is the obvious
extension of the usual capture-avoiding textual substitution, and the
rules of Figure~\ref{fig:terms} ensure that weakening is respected.

@@

respectively. Our calculus can similarly define a
$\mathtt{reverseBush}$ function that reverses the data in a bush. This

@@

\noindent
Direct calculation further reveals that term interpretations also
satisfy
\[\dsem{\Gamma; \Phi~|~\Delta \vdash
  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \dsem{\Gamma; \Phi~|~\Delta \vdash
  t [\ol{\alpha := K}][x := s]}\]
       {\color{red} Add if true:
$\dsem{\Gamma; \emptyset~|~\Delta \vdash L_{\ol{\alpha}} x.t_{\ol{\alpha}} x :
   \Nat^{\ol{\alpha}}\, F\, G} =  \dsem{\Gamma; \emptyset~|~\Delta \vdash
   t : \Nat^{\ol{\alpha}}\, F\, G}$}

@@

In this section we show how Theorem~\ref{thm:at-gen} and its
consequences can be used to prove free theorems that go beyond mere
naturality.  We also show that we can extend short cut fusion for
lists~\cite{glp93} to nested types, thereby formally proving
correctness of the categorically inspired theorem from~\cite{jg10}.

@@

Let $ \vdash g : \Nat^\alpha \,\onet\,\alpha$, let $g^\set =
\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$ and $g^\rel =
\relsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(g^\set,g^\set) \;= \;
(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\onet\,\alpha}\rho$ $ = \{\eta : K_1 \Rightarrow
\id\}$ $ = \{(\eta_1 : K_1 \Rightarrow \id, \eta_2 : K_1 \Rightarrow
\id)\}$ That is, $g^\set$ is a natural transformation from the
constantly $1$-valued functor to the identity functor in $\set$. In
particular, for every $S : \set$, $g^\set_S : 1 \to S$. Note, however,
that if $S = \emptyset$, then there can be no such morphism, so no
such natural transformation, and thus no term $\vdash g : \Nat^\alpha
\onet \,\alpha$, can exist.  That is, our calculus admits no
(non-terminating) terms with the closed type $\Nat^\alpha \onet
\,\alpha$ of the polymorphic bottom.

@@

Let $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$, \/$g^\set =
\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$, and $g^\rel =
\relsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  By
Theorem~\ref{thm:at-gen}, $(g^\set(\pi_1\rho),g^\set(\pi_2\rho)) =
g^\rel\rho$. Thus, for all $\rho \in \relenv$ and any $(a, b) \in
\relsem{\vdash \emptyset}\rho = 1$, eliding the only possible
instantiations of $a$ and $b$ gives that $(g^\set, g^\set) \; = \;
(g^\set(\pi_1 \rho), g^\set (\pi_2 \rho))$ $ \in \relsem{\vdash
  \Nat^\alpha \,\alpha\,\alpha}\rho$ $ = \{\eta : \id \Rightarrow
\id\}$ $ = \{(\eta_1 : \id \Rightarrow \id, \eta_2 : \id \Rightarrow
\id)\}$ That is, $g^\set$ is a natural transformation from the
identity functor on $\set$ to itself. Now let $S$ be any set.  For any
element $a$ of $S$, if $K_a :S \to S$ is the constantly $a$-valued
morphism on $S$ then instantiating the naturality square implied by
the above equality gives that $g^\set_S \circ K_a = K_a \circ
g^\set_S$, i.e., $g^\set_S \, a = a$, i.e., $g^\set_S = \id_S$.
Putting these two cases together we have that for every $S : \set$,
$g^\set_S = \id_S$, i.e., $g^\set$ is the identity natural
transformation for the identity functor on $\set$. So every closed
term $g$ of closed type $\Nat^\alpha\alpha\,\alpha$ always denotes the
identity natural transformation for the identity functor on $\set$,
i.e., every closed term $g$ of type $\Nat^\alpha\alpha\,\alpha$
denotes the polymorphic identity function.

@@

We conclude this subsection by showing, as promised above, that no
cutting down of set and relational interpretations of $\Lan$-types is
possible. We have:

@@

Having extended our calculus with both $\Lan$-types and terms of such
types, and having given the only sensible set and relational
interpretations for these types and terms, we now need to verify that
these interpretations give rise to a parametric model. The first step
in this process is to extend Lemma~\ref{lem:rel-transf-morph} to
$\Lan$-types by adding a clause for $\Lan$-types to the proof.
Unfortunately, however, Lemma~\ref{lem:rel-transf-morph} does not
extend to arbitrary $\Lan$-types, as the following example shows.

@@

Even when extending along constant-free polynomials, it is unclear how
to prove the Identity Extension Lemma for $\Lan$-types or, indeed,
whether it holds at all in their presence. To prove the IEL for
$\Nat$-types, for example, it is necessary to cut down the semantic
interpretation by requiring that the natural transformations in the
set interpretation preserve equalities. The impossibility of
restricting the semantic interpretations of $\Lan$-types, as shown in
Proposition~\ref{prop:full-lan}, suggests that the IEL may fail for
them. We have been able neither to prove nor disprove it thus far, but
our substantial and lengthy investigations into the issue lead us to
conjecture that it is not possible to define a functorial semantics
for $\Lan$-types that gives rise to parametric models for languages
supporting even constant-free polynomial primitive GADTs.

>>>>

Since the type systems of F$_\omega$ and dependent type theories are
designed to extend System F with far more than non-ADT data types, it
seems like serious overkill to pass to their parametric models to
reason about nested types in System F. Indeed, such calculi support
fundamentally new features --- e.g., the full hierarchy of
higher-kinds or term-indexed types --- adding complexity to their
models that is entirely unnecessary for reasoning about nested types.
In addition, to our knowledge no such models represent data types as
primitive fixpoints, so they cannot be used to obtain the results of
this paper.  (See the past paragraph of this section for a more
detailed discussion in the case of F$_\omega$.) Whether or not these
results even hold for such calculi is an open question.

@@

\[\begin{array}{cc}
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \zerot$}
\DisplayProof
&
\AXC{\phantom{$\Gamma,\Phi$}}
\UIC{$\Gamma;\Phi \vdash \onet$}
\DisplayProof
\end{array}\]\\[3ex]
\[\begin{array}{c}
\AXC{$\Gamma;\ol{\alpha^0} \vdash F$}
\AXC{$\Gamma;\ol{\alpha^0}  \vdash G$}
\BIC{$\Gamma;\Phi \vdash \Nat^{\ol{\alpha^0}}F \,G$}
\DisplayProof
\\[5ex]
\AXC{$\phi^k \in \Gamma \cup \Phi$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash F}$}
\BIC{$\Gamma;\Phi \vdash \phi^k \ol{F}$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\ol{\alpha^0},\phi^k \vdash F$}
\AXC{$\quad\quad\ol{\Gamma;\Phi \vdash G}$}
\BIC{$\Gamma;\Phi \vdash (\mu \phi^k.\lambda
  \ol{\alpha^0}. \,F)\,\ol{G}$}  
\DisplayProof
\\[6ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F + G$}
\DisplayProof
\\[5ex]
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi \vdash G$}
\BIC{$\Gamma; \Phi \vdash F \times G$}
\DisplayProof\\[2ex]
\end{array}\]
\end{defi}
We write $\vdash F$ for $\emptyset;\emptyset \vdash F$.
Definition~\ref{def:wftypes} ensures that the expected weakening rules
for well-formed types hold, i.e., if $\Gamma; \Phi \vdash F$ is
well-formed, then both $\Gamma, \ol{\psi}; \Phi \vdash F$ and $\Gamma;
\Phi, \ol{\phi} \vdash F$ are also well-formed. Note that weakening
does not change the contexts in which types can be formed.  For
example, a $\mathsf{Nat}$-type --- i.e., a type of the form
$\Nat^{\ol{\alpha}} F \,G$ --- can be formed in any functorial
context. But since $\Nat$-types contain no functorial variables, we
will form them in empty functorial contexts below whenever
convenient. Of course, we cannot {\em first} weaken the functorial
contexts of $F$ and $G$ to $\Gamma; \ol{\alpha}, \beta \vdash F$ and
$\Gamma; \ol{\alpha}, \beta \vdash G$, and {\em then} form
$\Nat^{\ol{\alpha}} F \,G$ in the weakened context $\Gamma; \beta$.
If $\Gamma;\emptyset \vdash F$ and $\Gamma;\emptyset \vdash G$, then
our rules allow formation of the type $\Gamma;\emptyset \vdash
\Nat^\emptyset F \,G$, which represents the System F type $\Gamma
\vdash F \to G$. Similarly, if $\Gamma;\ol{\alpha} \vdash F$, then our
rules allow formation of the type $\Gamma; \emptyset \vdash
\Nat^{\ol\alpha} \,\onet \,F$, which represents the System F type
$\Gamma; \emptyset \vdash \forall \ol\alpha . F$. However, some System
F types, such as $\forall \alpha. (\alpha \to \alpha) \to \alpha$, are
not representable in our calculus. Note that, in the rule for
$\mu$-types, no variables in $\Phi$ appear in the body $F$ of $\mu
\phi. \lambda \ol\alpha.F$. This will be critical to proving the
Identity Extension Lemma for our calculus.

@@

\begin{defi}\label{def:second-order-subst}
If \,$\Gamma; \Phi,\phi^k \vdash H$ and $\Gamma;\Phi, \ol{\alpha}
\vdash F$ with $|\ol\alpha| = k$, then the operation $H[\phi :=_{\ol
    \alpha} F]$ of {\em second-order type substitution along
  $\ol\alpha$} is defined by induction on $H$ as follows:
\[\begin{array}{lll}
\zerot[\phi :=_{\ol{\alpha}} F] & = & \zerot\\[0.5ex]
\onet[\phi :=_{\ol{\alpha}} F] & = & \onet\\[0.25ex]
(\Nat^{\ol\beta} G \,K)[\phi :=_{\ol{\alpha}} F]
& = & \Nat^{\ol\beta}\, G \,K\\
(\psi\ol{G})[\phi :=_{\ol{\alpha}} F] & = &
\left\{\begin{array}{ll}
\psi \,\ol{G[\phi :=_{\ol{\alpha}} F]} & \mbox{if } \psi \not = \phi\\
  F[\ol{\alpha  := G[\phi :=_{\ol{\alpha}} F]}] 
  & \mbox{if } \psi = \phi
\end{array}\right.\\[2.8ex]
(G + K)[\phi :=_{\ol{\alpha}} F] & = & G[\phi
  :=_{\ol{\alpha}} F] + K[\phi :=_{\ol{\alpha}} F]\\[0.5ex] 
(G \times K)[\phi :=_{\ol{\alpha}} F] & = &
G[\phi :=_{\ol{\alpha}} F] \times K[\phi
  :=_{\ol{\alpha}} F]\\[0.5ex]   
((\mu \psi. \lambda \ol{\beta}. G)\ol{K})[\phi :=_{\ol{\alpha}}
  F] & = & (\mu \psi. \lambda \ol{\beta}. \,G)\, \ol{K[\phi :=_{\ol{\alpha}} F]}
\end{array}\]
\end{defi}
\noindent
The idea is that the arguments to $\phi$ get substituted for the
variables in $\ol\alpha$ in each $F$ replacing an occurrence of
$\phi$. It is not hard to see that $\Gamma;\Phi \vdash H[\phi
  :=_{\ol{\alpha}} F]$.

@@

       \[\begin{array}{c}
       \AXC{$\Gamma; \ol{\alpha} \vdash F$}
       \AXC{$\Gamma; \ol{\alpha} \vdash G$}
       \AXC{$\Gamma; \ol{\alpha} \,|\, \Delta, x : F \vdash t: G$} 
       \TIC{$\Gamma; \Phi
         \,|\, \Delta \vdash L_{\ol{\alpha}} x.t : \Nat^{\ol{\alpha}} \,F \,G$}
       \DisplayProof\vspace*{-0.05in}
       \\\\
       \AXC{$\ol{\Gamma;\Phi \vdash K}$}
       \AXC{$\Gamma; \emptyset
         \,|\, \Delta \vdash t : \Nat^{\ol{\alpha}} \,F \,G$}
       \AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\alpha := K}]$}
       \TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
         G[\overline{\alpha := K}]$}
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \ol{\phi}, \ol{\gamma} \vdash H$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash F}$}
       \AXC{$\ol{\Gamma; \ol{\beta},\ol{\gamma} \vdash
           G}$}
       \TIC{$\Gamma; \Phi         ~|~\Delta
         \vdash \map^{\ol{F},\ol{G}}_H :
         \Nat^\emptyset\;(\ol{\Nat^{\ol{\beta},\ol{\gamma}}\,F\,G})\;
         (\Nat^{\ol{\gamma}}\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
             :=_{\ol{\beta}} G}])$} 
       \DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi, \ol{\alpha} \vdash H$}
       \UIC{$\Gamma; \Phi  \,|\, \Delta \vdash \tin_H :
         \Nat^{\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
           \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
         \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \\\\
       \AXC{$\Gamma; \phi,\ol{\alpha} \vdash H$}
       \AXC{$\Gamma; \ol\beta \vdash F$}
       \BIC{$\Gamma; \Phi  \,|\, \Delta \vdash \fold^F_H :
         \Nat^\emptyset\; (\Nat^{\ol{\beta}}\,H[\phi
           :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\;
         (\Nat^{\ol{\beta}}\,(\mu \phi.\lambda
         \ol{\alpha}.H)\ol{\beta}\,F)$} 
       \hspace*{0.2in}\DisplayProof\vspace*{-0.1in}
       \end{array}\]

@@

To define our term calculus we assume an infinite set $\cal V$ of term
variables disjoint from $\tvars$ and $\fvars$. If $\Gamma$ is a type
constructor context and $\Phi$ is a functorial context, then a {\em
  term context for $\Gamma$ and $\Phi$} is a finite set of bindings of
the form $x : F$, where $x \in {\cal V}$ and $\Gamma; \Phi \vdash
F$. We adopt the above conventions for denoting disjoint unions and
vectors in term contexts. If $\Delta$ is a term context for $\Gamma$
and $\Phi$, then the formation rules for the set of {\em well-formed
  terms over $\Delta$} are as in Figure~\ref{fig:terms}. In the rule
there for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all occurrences
of the type variables in $\ol{\alpha}$ in the types of $x$ and $t$, as
well as all occurrences of $x$ in $t$. In the rule for $t_{\ol K} s$
there is one type in $\ol K$ for every functorial variable in $\ol
\alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$ there is one type
$F$ and one type $G$ for each functorial variable in
$\ol\phi$. Moreover, for each $\phi^k$ in $\ol\phi$ the number of
functorial variables in $\ol\beta$ in the judgments for its
corresponding type $F$ and $G$ is $k$. In the rules for $\tin_H$ and
$\fold^F_H$, the functorial variables in $\ol{\beta}$ are fresh with
respect to $H$, and there is one $\beta$ for every
$\alpha$. Substitution for terms is the obvious extension of the usual
capture-avoiding textual substitution, and the rules of
Figure~\ref{fig:terms} ensure that weakening is respected.  Below we
form $L$-terms, which contain no free functorial variables, in empty
functorial contexts whenever convenient. We similarly form $\map$-,
$\tin$-, and $\fold$-terms, which contain no free functorial variables
or free term variables, in empty functorial contexts and empty term
contexts whenever convenient.

@@

respectively. Our calculus can similarly define a $\mathtt{rustle}$
function that changes the placement of the data in a bush. This

@@

\noindent
Direct calculation further reveals that term interpretations also
satisfy
\[\dsem{\Gamma; \Phi~|~\Delta \vdash
  (L_{\ol{\alpha}}x.t)_{\ol{K}}s} = \dsem{\Gamma; \Phi~|~\Delta \vdash
  t [\ol{\alpha := K}][x := s]}\]
and
\[\dsem{\Gamma; \emptyset~|~\Delta \vdash L_{\ol{\alpha}} x.t_{\ol{\alpha}} x :
   \Nat^{\ol{\alpha}}\, F\, G} = \dsem{\Gamma; \emptyset~|~\Delta
  \vdash t : \Nat^{\ol{\alpha}}\, F\, G}\]

@@

In this section we show how Theorem~\ref{thm:at-gen} and its
consequences can be used to prove free theorems. Those in
Sections~\ref{sec:ft-adt},~\ref{sec:short-cut},
and~\ref{sec:short-cut-nested} go beyond mere naturality.  We also
show that we can extend short cut fusion for lists~\cite{glp93} to
nested types, thereby formally proving correctness of the
categorically inspired theorem from~\cite{jg10}.

@@

Suppose $ \vdash g : \Nat^\alpha \,\onet\,\alpha$ and $g^\set =
\setsem{\vdash g : \Nat^\alpha \,\onet\,\alpha}$. Then $g^\set$ is a
natural transformation from the constantly $1$-valued functor to the
identity functor in $\set$.  In particular, for every $S : \set$,
$g^\set_S : 1 \to S$. Note, however, that if $S = \emptyset$, then
there can be no such morphism, so no such natural transformation, and
thus no term $\vdash g : \Nat^\alpha \onet \,\alpha$, can exist.  That
is, our calculus admits no (non-terminating) terms with the closed
type $\Nat^\alpha \onet \,\alpha$ of the polymorphic bottom.

@@

Suppose $ \vdash g : \Nat^\alpha \,\alpha\,\alpha$ and $g^\set =
\setsem{\vdash g : \Nat^\alpha \,\alpha\,\alpha}$.  Then $g^\set$ is a
natural transformation from the identity functor on $\set$ to
itself. If $S$ is any set, if $a$ is any element of $S$, and if $K_a
:S \to S$ is the constantly $a$-valued morphism on $S$, then
naturality of $g^\set$ gives that $g^\set_S \circ K_a = K_a \circ
g^\set_S$, i.e., $g^\set_S \, a = a$, i.e., $g^\set_S = \id_S$.  That
is, $g^\set$ is the identity natural transformation for the identity
functor on $\set$. So every closed term $g$ of closed type
$\Nat^\alpha\alpha\,\alpha$ always denotes the identity natural
transformation for the identity functor on $\set$, i.e., every closed
term $g$ of type $\Nat^\alpha\alpha\,\alpha$ denotes the polymorphic
identity function.

@@

We conclude this subsection by showing, as promised above, that no
cutting down of set and relational interpretations of $\Lan$-types by
taking subsets is possible. We have:

@@

Having extended our calculus with both $\Lan$-types and terms of such
types, and having given sensible set and relational interpretations
for these types and terms, we now need to verify that these
interpretations give rise to a parametric model. The first step in
this process is to extend Lemma~\ref{lem:rel-transf-morph} to
$\Lan$-types by adding a clause for $\Lan$-types to the proof.
Unfortunately, however, Lemma~\ref{lem:rel-transf-morph} does not
extend to arbitrary $\Lan$-types, as the following example shows.

@@

Even when extending along constant-free polynomials, it is unclear how
to prove the Identity Extension Lemma for $\Lan$-types or, indeed,
whether it holds at all in their presence. To prove the IEL for
$\Nat$-types, for example, it is necessary to cut down the semantic
interpretation by requiring that the natural transformations in the
set interpretation preserve equalities. The impossibility of
restricting the semantic interpretations of $\Lan$-types, as shown in
Proposition~\ref{prop:full-lan}, suggests that the IEL may fail for
them. We have been able neither to prove nor disprove it thus far, but
our substantial and lengthy investigations into the issue lead us to
suspect that it may not actually be possible to define a functorial
semantics for $\Lan$-types that gives rise to parametric models for
languages supporting even constant-free polynomial primitive GADTs.
